---
title: "Submission for Course5 Week1"
author: "Arati"
date: "December 10, 2018"
output: html_document
---

## 1. Loading data 

Loading activity.csv into a dataframe and looking at contents:

```{r, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, fig.path='figure/')
setwd("C:/Users/Arati/Documents/R_studying/Course5/repdata_data_activity")
ac <- read.csv("activity.csv")
sapply(ac,class)
head(ac)
```

##2.  What is mean total number of steps taken per day?

```{r, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

### Calculate total number of steps per day

```{r, include=TRUE, results=TRUE}
steps_byday <- ac %>% group_by(date) %>% summarize(steps_count = sum(steps,na.rm=TRUE))
head(steps_byday)
```

### Plotting histogram of total steps per day

```{r, include=TRUE, results=TRUE}
with(steps_byday,hist(steps_count,breaks=5,col="red",main="Histogram of total steps by day"))
abline(v=mean(steps_byday$steps_count,na.rm=TRUE),col="blue",lwd=3)
abline(v=median(steps_byday$steps_count,na.rm=TRUE),col="green",lwd=3)
rug(steps_byday$steps_count)
legend("topright",legend=c("Mean","Median"),col=c("blue","green"),lwd=3)
```


### Calculating mean and median total steps per day

```{r, include=TRUE, results=TRUE}
mean_steps_ignorena <- mean(steps_byday$steps_count)
median_steps_ignorena <- median(steps_byday$steps_count)
mean_steps_removena <- mean(steps_byday$steps_count,na.rm=TRUE)
median_steps_removeena <- median(steps_byday$steps_count,na.rm=TRUE)
print(mean_steps_ignorena)
print(mean_steps_removena)
print(median_steps_ignorena)
print(median_steps_removeena)
```

##2. What is the average daily activity pattern?

### Calculating the mean steps per interval over all days
```{r,inclue=TRUE,results=TRUE}
mnsteps <- ac %>% group_by(interval) %>% summarize(meansteps=mean(steps,na.rm=TRUE))
head(mnsteps)
```

### Plotting time series plot of mean steps per interval in a day

```{r,include=TRUE, results=TRUE}
h <- mnsteps$interval%/%100
m <- mnsteps$interval%%100
mnsteps$timestamp <- strptime(as.character(paste(h,m)),"%H%M")
with(mnsteps,plot(timestamp,meansteps,type="l"))
```

### Finding time interval at which the average steps taken was highest

```{r, include=TRUE,results=TRUE}
maxsteps <- max(mnsteps$meansteps)
maxstepstime <- mnsteps$timestamp[which(mnsteps$meansteps == maxsteps)]
print(maxsteps)
print(strftime(maxstepstime,"%H-%M"))
```

## 3. Imputing missing values

### Total number of missing steps values

```{r,include=TRUE, results=TRUE}
n_nasteps <- length(which(is.na(ac$steps)))
print(n_nasteps)
```

### Strategy for imputing missing values:
Strategy one: impute missing values with the following formulae:
1. Mean steps for that day
2. Mean steps for that interval over all days

The above three strategies will be compared by computing the total steps per day and comparing with the missing value data set.

1. Mean steps for that day
```{r, include = TRUE, results = TRUE}
ac1 <- ac
ind <- which(is.na(ac1$steps))
ac1$steps1 <- ac1$steps
for (i in ind) {
  datelookupvalue <- ac1$date[i]
  datematchindex <- which(steps_byday$date==datelookupvalue)
  ac1$steps1[i] <- steps_byday$steps_count[datematchindex]
}
head(ac1)
```

2. Computing mean values for interval over all days
```{r, include = TRUE, results = TRUE}
ac2 <- ac
head(ac2)
ac2$steps2 <- ac2$steps
for (i in ind) {
  intervallookupvalue <- ac2$interval[i]
  intervalmatchindex <- which(mnsteps$interval==intervallookupvalue)
  ac2$steps2[i] <- mnsteps$meansteps[intervalmatchindex]
}
head(ac2)
```

### Comparing the mean, median and histograms for the two imputing strategies.
#### Strategy 1. Mean steps for that day
```{r, include=TRUE, results=TRUE}
steps_byday1 <- ac1 %>% group_by(date) %>% summarize(steps_count1 = sum(steps1))
head(steps_byday1)
meansteps1 <- mean(steps_byday1$steps_count1)
mediansteps1 <- median(steps_byday1$steps_count1)
hist(steps_byday$steps_count,col=rgb(1,0,0,0.5),ylim=c(0,30))
hist(steps_byday1$steps_count1,col=rgb(0,1,0,0.5),add=T)
legend("topright",legend=c("no imputation","imputed mean steps for the day"),col=c(rgb(1,0,0),rgb(0,1,0)),lwd=3)
abline(v=meansteps1,col="green",lwd=3)
abline(v=mean_steps_ignorena,col="red",lwd=3)
abline(v=mediansteps1,col="green",lty=2,lwd=3)
abline(v=median_steps_ignorena,col="red",lty=2,lwd=3)
print(meansteps1)
print(mean_steps_ignorena)
print(mediansteps1)
print(median_steps_ignorena)
```

Strategy 1 (i.e imputing missing data with the mean of the day) predictably gives you the same mean and median values of total steps per day.
The time-series plot might well look different. 

#### Strategy2. Mean steps for that interval
```{r, include=TRUE, results=TRUE}
steps_byday2 <- ac2 %>% group_by(date) %>% summarize(steps_count2 = sum(steps2))
head(steps_byday2)
meansteps2 <- mean(steps_byday2$steps_count2)
mediansteps2 <- median(steps_byday2$steps_count2)
hist(steps_byday$steps_count,col=rgb(1,0,0,0.5),ylim=c(0,40))
hist(steps_byday2$steps_count2,col=rgb(0,1,0,0.5),add=T)
legend("right",legend=c("no imputation","imputed mean steps for the interval"),col=c(rgb(1,0,0),rgb(0,1,0)),lwd=3,cex=.8)
abline(v=meansteps2,col="green",lwd=1)
abline(v=mean_steps_ignorena,col="red",lwd=3)
abline(v=mediansteps2,col="green",lty=2,lwd=1)
abline(v=median_steps_ignorena,col="red",lty=2,lwd=3)
print(meansteps2)
print(mean_steps_ignorena)
print(mediansteps2)
print(median_steps_ignorena)
```

Strategy 2 has changed the distribution of steps per day somewhat - makes it look more normal. 
The mean value has shifted to the right (dashed line is for imputed data, solid line is without imputation).
Also of interest: the mean and median values for the imputed data are the same. 

##4.  Are there differences in activity patterns between weekdays and weekends?

Strategy1 will be tested for time series plots by weekday/weekend

```{r, include=TRUE, results=TRUE}
library(lubridate)
ac1$date1 <- as.POSIXct(strptime(ac1$date,"%Y-%m-%d"))
ac1$weeknum <- wday(ac1$date)
ac1$weektype <- factor(ac1$weeknum,levels = 1:7, labels = c("weekend",rep("weekday",length=5),"weekend"))

mnsteps1f  <- ac1 %>% group_by(weektype,interval) %>% summarize(meansteps1 = mean(steps1))
print(mnsteps1f[1:5,])
library(lattice)
xyplot(mnsteps1f$meansteps1 ~ mnsteps1f$interval|mnsteps1f$weektype,type="l",layout=c(1,2))